---
title: "Final Homework"
author: "Kyle Emerson"
date: "7/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = TRUE)
```
   Albasini Dam in South Africa has been hypothesized to be experiencing a decrease in precipitation over the last 4 decades. To evaluate this hypothesis, daily precipitation records taken from the Albasini Dam weather station have been accumulated with the purpose of statistically anyalyzing any trends of precipitation over this time.
   
   We evaluated the annual precipitation over the course of 1980-2020 and found despite a large outlier of precipitation in 2000, no discernable trend was discovered (Fig. 1). We then compiled the average precipitation per month over the duration of 1980-2020 and saw the largest proportion of precipitation occured over the summer months (Fig. 2). To evaluate the significance of our data, we conducted a Two Sample t-test to compare the annual precipitation during two time intervals: 1980-2000 and 2001-2020. The results of this t-test found that these two time interval samples were not statistically significant from each other (t = 0.32064, p = 0.7493). Despite this non-significant result, we did evaluate a slight decrease in annual precipitation in the 2001-2020 time interval, signalling that there may be a trend of decreasing precipitation in the area (Fig. 3). 
   
   Despite not finding any statistcally significant data to support our claim that precipitation is decreasing in this area, further investigation is required to evaluate precipitation trends. Further studies could widen the time parameters (i.e. start evaluation from 1960-2020) while still evaluating hydrologic years. Further, decreasing the width of our time intervals from 20 to 5 years could yield more information regarding recent precipitation trends. 
   
   



## Precipitation Patterns  
*Albasini Dam, South Africa*  
```{r include=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(forcats)

# the data are provided via URL, the "filename" will not need adjustment - this may not apply to the options of the command.  If you prefer to work locally, the dataset is in the GitHub repository and the Box folder.  For the purposes of this homework, you may ignore the quality codes (in fact, you need to in order to process all of the data).  
prcp <- read_csv("https://duq.box.com/shared/static/d1523vk9rnilrtsuuhodffwj6vjygywk.csv", skip = 9)
prcp <- prcp[1:14610,]

# process dates with lubridate.  For help, see the lubridate cheat sheet in the Box folder.
prcp$dt <- ymd(prcp$DATE)
```

Data are analyzed based on hydrologic year, which South Africa runs from July to August.

```{r include=FALSE}
# Pull year and month from dt
prcp$year <- year(prcp$dt)
prcp$mont <- month(prcp$dt)
# Preallocate "hydro.y" for hydrologic year and hydro.m
prcp$hydro.y <- prcp$year
prcp$hydro.m <- prcp$mont
# DEFINE: hydrologic year is the year that the hydrologic year ENDS.  Ex: Hydrologic year 2020 begins August 2019 and goes to July 2020.
for (i in 1:nrow(prcp)) {
      if (prcp$mont[i] >= 8) {
            prcp$hydro.y[i] <- prcp$hydro.y[i] + 1 # adds one year if the month is August or later.
            prcp$hydro.m[i] <- prcp$hydro.m[i] - 7 # pulls month back so that August(8) is month 1
      } else {
            prcp$hydro.m[i] <- prcp$hydro.m[i] + 5 # pushes month forward so January(1) is month 6
      }
}
```

### Annual Precipitation

```{r echo=FALSE}
annual <- prcp %>%
       filter(hydro.y >= 1980)%>%
      group_by(hydro.y) %>%  
      summarize(total = sum(`DAILY RAIN`, na.rm = TRUE))
ggplot(annual, aes(x = hydro.y, y = total) ) +
      geom_bar(stat = "identity", fill = "steelblue") +
      ggtitle ("Annual Precipitation at Albasini Dam, South Africa") +
      theme(plot.title = element_text(face = "bold", size = (13), colour = "steelblue4")) +
      labs(x = "Year", y = "Annual Precipitation (mm)") +
      theme(panel.background = element_rect(fill = "white", colour = "black")) +
      theme(aspect.ratio = 1) +
      theme(axis.text = element_text(face = "plain", size = 11)) +
      theme(axis.title = element_text(face = "bold", size = 12))  
```

Figure 1: Annual precipitation over the time period of 1980-2020. Annual precipitation is presented in millimeters and was obtained from the Albasini Dam weather station in South Africa. Years are represented as hydrologic years, which run from August to July in South Africa.

### Typical Monthly Precipitation

```{r echo=FALSE}
prcp$hydro.ym <- 100 * prcp$hydro.y + prcp$hydro.m
monthly <- prcp %>%
      group_by(hydro.ym) %>%
      summarize(total = sum(`DAILY RAIN`, na.rm = TRUE))

mon_pattern <- monthly %>%
      mutate(year = floor(hydro.ym/100)) %>%
      mutate(mon = hydro.ym - (100 * floor(hydro.ym/100))) %>%
      group_by(mon) %>%
      summarize(prcp = mean(total), std = sd(total))
mon_pattern$m <- factor(c("Aug", "Sep", "Oct", "Nov", "Dec", "Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul")) # this puts the months in this order; otherwise, alphabetical is used.

ggplot(mon_pattern, aes(x = fct_reorder(m, mon), y = prcp)) + # fct_reorder is a command from forcats library that orders the factor, m (the name of the month)
      geom_bar(stat = "identity", fill = "steelblue") +
      geom_errorbar(aes(ymin = prcp-std, ymax = prcp+std), width = .1, position = position_dodge(.9)) +
      ggtitle ("Average Monthly Precipitation at Albasini Dam, South Africa") +
      theme(plot.title = element_text(face = "bold", size = (13), colour = "steelblue4")) +
      labs(x = "Month" , y = "Average Monthly Precipitation (mm)" ) +
      theme(panel.background = element_rect(fill = "white", colour = "black")) +
      theme(aspect.ratio = 1) +
      theme(axis.text = element_text(face = "plain", size = 11)) +
   theme(axis.title = element_text(face = "bold", size = 12))  


```

Figure 2: Average precipitation per month over the time period of 1980-2020. Average precipitation is presented in millimeters and was obtained from the Albasini Dam weather station in South Africa. Years are represented as hydrologic years, which run from August to July in South Africa. Error bars were calculated as the means +/- standard deviation.

### T-test for the Scores 1980-2000 compared to 2001-2020

```{r echo=FALSE}
# t.test
prcp1980 <- prcp %>%
      filter(hydro.y <= 2000) %>%
      group_by(hydro.y) %>%
      summarize(total = sum(`DAILY RAIN`, na.rm = TRUE))
prcp2000 <- prcp %>%
      filter(hydro.y >= 2001 ) %>%
      group_by(hydro.y) %>%
      summarize(total = sum(`DAILY RAIN`, na.rm = TRUE))
t <- t.test(prcp1980, prcp2000)
annual$Interval <- as.character(annual$hydro.y) # preallocation
for (i in 1:(nrow(annual)))  {
      if (annual$hydro.y[i] <= 2000) {
            annual$Interval[i] <- "1981-2000"
      } else {
            annual$Interval[i] <- "2001-2020"
      }
}
ggplot(annual, aes(x= Interval, y = total, fill = Interval)) +
      geom_boxplot() +
      stat_summary(fun = "mean", geom = "point", shape = 8, size = 2, color = "white") +
      labs(x = "Year", y = "Annual Precipitation (mm)") +
      theme(panel.background = element_rect(fill = "white", colour = "black")) +
      theme(aspect.ratio = 1) +
      ggtitle("Annual Precipitation at Albasini Dam, South Africa") +
      theme(plot.title = element_text(face = "bold", size = (13), colour = "steelblue4")) +
      theme(axis.text = element_text(face = "bold", size = 11))  +
      theme(axis.title = element_text(face = "bold", size = 12))  

```

Figure 3: Annual precipitation over the course of 1981-2000 and 2001-2020. Annual precipitation is presented in millimeters and was calculated by averaging the annual precipitation of every year within each time interval. Data was obtained from the Albasini Dam weather station in South Africa. Years are represented as hydrologic years, which run from August to July in South Africa. The t-test shows p = 0.7493. This suggests that these two populations are not statistically different.  The trend shown by the boxplot is consistent with the overall decrease in precipitation in the region. White asterisks represent the mean of the data. In each box plot, the center line represents the median, the length of the box extends through the IQR, and whiskers extend to 1.5x the IQR. All points outsde this range are plotted individually.

